---
title: "Benchmarking"
author: "Abhijit Dasgupta"
date: "10/26/2016"
output: html_document
---

```{r setup, echo=F}
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      warning=F,
                      message = F,
                      comment = '> ')
```
```{r, message=F, warning=F}
# Loading packages we will need in this exercise

library(rbenchmark) # Benchmarking package
library(readr) # Hadley's package for reading data
library(reshape2)
library(dplyr)
library(tidyr) # Hadley's package for reshaping tidy data
library(data.table) # an alternative package by Matt Dowle
                    # for storing and manipulating data. It 
                    # stores data in a data.table object rather
                    # than a data.frame object
```

## Introduction

We are going to repeat our exercise of getting the p-values of multiple t-tests from an 
expression data set as we did in [Lecture 6](../lecture6.html). However, now, we will use the full
dataset, which has 12553 proteins, rather than just the 10 we were looking at before. So we are going to get 12553 p-values by doing 12553 t-tests comparing expression between ER-positive and ER-negative patients. 

## Reading the data into R

This data set is stored as a compressed (zipped) file called _fullExpression.zip_, which is 3.6Mb. The actual data is not huge, but is still 13Mb when stored as a text (csv) file. 

First of all, we actually don't need to uncompress this file to use the dataset. R has functions that will help
us get the data out of a zipped file. First, let's see what is in the zipped file.
```{r}
unzip('../lecture6_data/fullExpression.zip', list=T) # just lists files
```
We see that this file only has one csv file named *BreastCancer_Expression_full.csv*. We're going to read this file in a few ways. First, let's use the way we've learnt, that comes with R. 
```{r}
datafile <- unz('../lecture6_data/fullExpression.zip',
                'BreastCancer_Expression_full.csv')
expression_dat <- read.csv(datafile,
                           stringsAsFactors=F)
```
Next we will use the `read_csv` function from the `readr` package
```{r}
datafile <- unz('../lecture6_data/fullExpression.zip',
                'BreastCancer_Expression_full.csv')
expression_dat <- read_csv(datafile) # from readr
```
Third, we will use the `fread` function from the `data.table` package. Unfortunately, this function cannot
read zipped files directly, so we'll have to unzip the file first. The `data.table` package works with data in objects called `data.table` rather than the usual `data.frame` object. There are some advantages to this package, like speed, but I personally find the syntax harder. However, the `fread` function is fantastic!!!
```{r}
if(!file.exists('BreastCancer_Expression_full.csv')){ # Check if the file already exists
  unzip('../lecture6_data/fullExpression.zip',
        'BreastCancer_Expression_full.csv', # Specify the file to be extracted
        exdir = '.') # Specify the directory where it will be extracted
}
expression_dat <- fread('BreastCancer_Expression_full.csv') # fread creates a data.table
expression_dat <- as.data.frame(expression_dat) # we convert it to a data.frame
```

You can check for yourself whether the data extracted using each of the three methods is the same. Let's find out which one is the fastest, though. 
```{r, results='as.is'}
library(rbenchmark)
benchmark_reading <- benchmark(baseR = read.csv('BreastCancer_Expression_full.csv', stringsAsFactors=F),
          readr = read_csv('BreastCancer_Expression_full.csv'),
          data.table = fread('BreastCancer_Expression_full.csv'),
          columns = c('test', 'elapsed','replications','relative'),
          order = 'elapsed',
          replications = 1)
knitr::kable(benchmark_reading, row.names = F)
```





