---
title: "Statistical Modeling"
author: "Abhijit Dasgupta"
date: "November 7, 2018"
output:
  xaringan::moon_reader:
    css: [default, './robot.css', './robot-fonts.css']
    #css: [default, metropolis, metropolis-fonts]
    nature:
      ratio: '16:9'
      highlightLanguage: R
      countIncrementalSlides: false
      highlightStyle: zenburn
      highlightLines: true

---

layout: true

<div class="my-header">
<span>BIOF 339, Fall 2018</span></div>

---

```{r setup, include=FALSE, message = F, warning = F}
knitr::opts_chunk$set(echo = FALSE, message = F, warning = F, comment="", cache=T)
library(tidyverse)
library(broom)
library(readxl)
```

# Goals today

.pull-left[
+ Using data models
    - Types and purpose
+ Common statistical models
    - Linear models
    - Logistic models
+ The "formula interface" in R to express models
]
.pull-right[
+ Learning to extract and use the results of models
    - The `broom` package
    - Graphical representations
+ What predictors/covariates are "important"
+ Model building and selection
+ Good practices
]

---
class: inverse, middle, center


.bigblockquote[All models are wrong, but some are useful]

.right[G.E.P. Box]

---

# Models

Models are our way of understanding nature, usually using some sort of mathematical expression

Famous mathematical models include Newton's second law of motion, the laws of thermodynamics, the ideal gas law

All probability distributions, like Gaussian, Binomial, Poisson, Gamma, are models

Mendel's laws are models **that result in** particular mathematical models for inheritance and population prevalence

---

# Models

We use models all the time to describe our understanding of different processes

+ Cause-and-effect relationships
--

+ Supply-demand curves
--

+ Financial planning
--

+ Optimizing travel plans (perhaps including traffic like _Google Maps_)
--

+ Understanding the effects of change
    - Climate change
    - Rule changes via Congress or companies
    - Effect of a drug on disease outcomes
    - Effect of education and behavioral patterns on future earnings
    
    
---

# Data-driven models

> Can we use data collected on various aspects of a particular context to understand the relationships between the different aspects?

+ How does increased smoking affect your risk of getting lung cancer? (causality/association)
    - Does genetics matter?
    - Does the kind of smoking matter?
    - Does gender matter?

---

# Data-driven models

> Can we use data collected on various aspects of a particular context to understand the relationships between the different aspects?

+ What is your lifetime risk of breast cancer? (prediction)
    - What if you have a sister with breast cancer?
    - What if you had early menarche?
    - What if you are of Ashkenazi Jewish heritage?
    
.right[[The Gail Model from NCI](https://bcrisktool.cancer.gov/)]

---

# Association models 

These are more traditional, highly interpretative models that look at **how** different predictors
affect outcome.

+ Linear regression
+ Logistic regression
+ Cox proportional hazards regression
+ Decision trees

Since these models have a particular known structure determined by the modeler, they can 
be used on relatively small datasets

You can easily understand which predictors have more "weight" in influencing the outcome

You can literally write down how a prediction would be made

---

# Predictive models

These are more recent models that primarily look to provide good predictions of an outcome, and 
the way the predictions are made is left opaque (often called a _black box_)

+ Deep Learning (or Neural Networks)
+ Random Forests
+ Support Vector Machines
+ Gradient Boosting Machines

These models require data to both determine the structure of the model as well as make the predictions, so they require lots of data to _train_ on

The relative "weight" of predictors in influencing the **predictions** can be obtained

The effect of individual predictors is not easily interpretable, though this is changing

They require a different **philosophic perspective** than traditional association models

---

# Datasets

We will use the `pbc` data from the `survival` package, and the in-built `mtcars` dataset.

```{r, echo=T}
library(survival)
data(pbc)
str(pbc)
```

---
class: inverse, middle, center

# The formula interface

---

# Representing model relationships

In R, there is a particularly convenient way to express models, where you have

- one dependent variable
- one or more independent variables, with possible transformations and interactions

```
y ~ x1 + x2 + x1:x2 + I(x3^2) + x4*x5
```

--

`y` depends on ...
--

- `x1` and `x2` linearly
--

- the interaction of `x1` and `x2` (represented as `x1:x2`)
--

- the square of `x3` (the `I()` notation ensures that the `^` symbol is interpreted correctly)
--

- `x4`, `x5` and their interaction (same as `x4 + x5 + x4:x5`)

---

# Representing model relationships

```
y ~ x1 + x2 + x1:x2 + I(x3^2) + x4*x5
```

This interpretation holds for the vast majority of statistical models in R

- For decision trees and random forests and neural networks, don't add interactions or transformations, since the model will try to figure those out on their own

---

# Our first model

```{r, echo=T}
myLinearModel <- lm(chol ~ bili, data = pbc)
```

Note that everything in R is an **object**, so you can store a model in a variable name.

This statement runs the model and stored the fitted model in `myLinearModel`

R does not interpret the model, evaluate the adequacy or appropriateness of the model, or comment on whether looking at the relationship between cholesterol and bilirubin makes any kind of sense.

--

.bigblockquote[It just fits the model it is given]

---

# Our first model

```{r, echo=T}
myLinearModel
```

> Not very informative, is it?

---

# Our first model

```{r, echo=T}
summary(myLinearModel)
```

> A little better

???

Talk about the different metrics in this slide


---

# Our first model

```{r, echo=T}
broom::tidy(myLinearModel)
```

```{r, echo=T}
broom::glance(myLinearModel)
```

---

# Our first model

We do need some sense as to how well this model fit the data

.pull-left[
```{r, echo=T, eval=F}
# install.packages('ggfortify')
library(ggfortify)
autoplot(myLinearModel)
```
]
.pull-right[
```{r, echo=F, eval=T}
# install.packages('ggfortify')
library(ggfortify)
autoplot(myLinearModel)
```
]

---

# Our first model

Let's see if we have some strangeness going on

--
.pull-left[
```{r, echo=T, eval=F}
ggplot(pbc, aes(x = bili))+geom_density()
```

We'd like this to be a bit more "Gaussian" for better behavior
]
.pull-right[
```{r, echo=F, eval=T}
ggplot(pbc, aes(x = bili))+geom_density()
```

]

---

# Our first model

Let's see if we have some strangeness going on

.pull-left[
```{r, echo=T, eval=F}
ggplot(pbc, aes(x = log(bili)))+geom_density()
```

]
.pull-right[
```{r, echo=F, eval=T}
ggplot(pbc, aes(x = log(bili)))+geom_density()
```

]

---

# Our first model

```{r, echo=T}
myLinearModel2 <- lm(chol~log(bili), data = pbc)
summary(myLinearModel2)
```

---

# Our first model

```{r}
autoplot(myLinearModel2)
```

---

# Just the residual plot, please

```{r, echo=T}
autoplot(myLinearModel2, which=1)
```

--

> OR

---

# Just the residual plot, please

```{r, echo=T}
d <- broom::augment(myLinearModel2)
d
```

---

# Just the residual plot, please

```{r, echo=T, fig.height=5}
ggplot(d, aes(x = .fitted, y = .resid))+geom_point()+ geom_smooth(se=F)+
  labs(x = 'Fitted values', y = 'Residual values')
```


---

# Predictions

```{r, echo=T}
head(predict(myLinearModel2, newdata = pbc))
```

The `newdata` has to have the same format and components as the original data the model was trained on

---

# Categorical predictors

```{r, echo=T}
myLM3 <- lm(chol ~ log(bili) + sex, data = pbc)
broom::tidy(myLM3)
```

R has a somewhat unfortunate notation for categorical varialbes here, as `{variable name}{level}`

---
class: inverse, middle, center

# Logistic regression

---

# The logistic transformation

For an outcome which is binary (0/1), what is really modeled is the **probability** that the outcome is 1, usually denoted by _p_.

However, we know $0 \leq p \leq 1$, so what if the model gives a prediction outside this range!!

The logistic transform takes _p_ to 

$$
\text{logit}(p) = \log\left(\frac{p}{1-p}\right)
$$

and we model _logit(p)_, which has a range from $-\infty$ to $\infty$

---

# Logistic regression

Logistic regression is a special case of a **generalized linear model**, so the function we use to run a logistic regression is `glm`

```{r, echo = T}
myLR <- glm(spiders ~ albumin + bili + chol, data = pbc, family = binomial)
myLR
```

- We have to add the `family = binomial` as an argument, since this is a special kind of GLM
- All these models only use complete data; they kick out rows with missing data

---

# Logistic regression

```{r, echo=T}
broom::tidy(myLR)
```
--

```{r, echo=T}
broom::glance(myLR)
```

---

# Predictions from logistic regression

```{r, echo=T}
head(predict(myLR))
```
--

These are on the "wrong" scale. We would expect probabilities
--

```{r, echo=T}
head(predict(myLR, type='response'))
```

--

or you can use `plogis(predict(myLR))` for the inverse logistic transform

---
class: inverse, middle, center

# Model selection

---

# How to get the "best" model

Generally getting to the best model involves

- looking at a lot of graphs
- Fitting lots of models
- Comparing the model fits to see what seems good

Sometimes if you have two models that fit about the same, you take the smaller, less complex model (Occam's Razor)

Generally it is not recommended that you use automated model selection methods. It screws up your error rates and may not be the right end result for your objectives

--

> Model building and selection is an art

---

# Clues to follow

You can look at the relative weights (size of coefficient and its p-value) of different predictors

- These weights will change once you change the model, so be aware of that

--

You can trim the number of variables based on collinearities

- If several variables are essentially measuring the same thing, use one of them

--

You can look at residuals for clues about transformations

--

You can look at graphs, as well as science, for clues about interactions (synergies and antagonisms)

---

# Automated model selection

```{r, echo=T}
# install.packages('leaps')
mtcars1 <- mtcars %>% mutate_at(vars(cyl, vs:carb), as.factor)
all_subsets <- regsubsets(mpg~., data = mtcars1)
all_subsets
```

---

# Automated model selection

Which has the best R<sup>2</sup>?

```{r, echo=T}
ind <- which.max(summary(all_subsets)$adjr2)
summary(all_subsets)$which[ind,]
```

