---
title: "Summarize outcomes into new variables"
author: "Abhijit Dasgupta"
date: 2017-11-22
categories: ["R"]
tags: ["Data munging"]
---
```{r, echo=F}
knitr::opts_chunk$set(echo=TRUE, eval=FALSE, collapse = TRUE)
```

## An example

Suppose you have a fairly detailed set of outcomes for cause of death. You want to summarize these
outcomes into categories and create new variables indicating whether a person's death could be 
attributed to a particular category or not. For example, suppose deaths from fentanyl are recorded as
one of:

+ Fentanyl Intoxication
+ Fentanyl and Morphine Intoxication
+ Combined Drug Intoxication (Heroin, Cocaine and Fentanyl)
+ Fentanyl-involved 

These are all possiblities in your dataset under a variable `Cause_of_death`. We will create a new
variable, `Fentanyl`, that is a boolean variable which is `TRUE` if the person's death is fentanyl-related 
and `FALSE` if it isn't.

## Approach 1

This is a general approach. First make a vector of all the possible ways fentanyl-related deaths are 
records:

```{r}
fentanyl_related <- c('Fentanyl Intoxication', 'Fentanyl and Morphine Intoxication',
                      'Combined Drug Intoxication (Heroin, Cocaine and Fentanyl)',
                      'Fentanyl-involved')
```
We then can create the new variable quite easily as
```{r}
dat <- dat %>% mutate(Fentanyl = Cause_of_death %in% fentanyl_related)
```

## Approach 2

This approach leverages the text processing capabilities of the `stringr` package. We note that all
the outcomes in `fentanyl_related` have the word "Fentanyl" in it. So we can create the `Fentanyl` 
variable using the following code:
```{r}
dat <- dat %>% 
  mutate(Fentanyl = stringr::str_detect(Cause_of_death, 'Fentanyl'))
```

We could also protect ourselves against possible recording errors (in terms of capitalization) by making
all the entries lower-case and then testing for the lower-case "fentanyl". This avoids the issues with
`ignore.case` that are required for the `grep` family of functions.
```{r}
dat <- dat %>% 
  mutate(Fentanyl = stringr::str_detect(tolower(Cause_of_death), 'fentanyl'))
```



