---
title: "Practice Assignment 1 Solution"
date: "September 19, 2016"
output: html_document
---


#### Useful functions

1. If you have a continuous variable `x`, you can compute its mean using `mean(x)`, its
standard deviation by `sd(x)` and its median by `median(x)`

2. If you have a categorical variable `y`, you can find out the frequency 
distribution of the different levels by `table(y)`

-------------

You can download a data set from a breast cancer study [here](clinical_data_breast_cancer_modified.csv) (right-click on the link to download). 
This data consists of several numerical and categorical variables, and some character variables. The variables are named in 
the data set. 
```{r}
dat <- read.csv('http://www.araastat.com/BIOF339_PracticalR/Lectures/lecture2_data/clinical_data_breast_cancer_modified.csv', 
                stringsAsFactors=FALSE)
```


Answer the following questions:

1. How many variables are in this data set?
```{r}
ncol(dat)
```

2. How many observations are in this data set?
```{r}
nrow(dat)
```

1 and 2 can also be answered with
```{r}
dim(dat)
```
The information is also available using 
```{r}
str(dat)
```


3. What is the mean and median age of the participants in this study?
```{r}
mean(dat[,'Age.at.Initial.Pathologic.Diagnosis'])
median(dat[,'Age.at.Initial.Pathologic.Diagnosis'])
```
Alternatively,
```{r}
mean(dat$Age.at.Initial.Pathologic.Diagnosis)
median(dat$Age.at.Initial.Pathologic.Diagnosis)
```


4. How many women are ER-positive, PR-positive and Her2-positive? (use the `table` function on appropriate variables)
```{r}
table(dat[,'ER.Status'])
table(dat[,'PR.Status'])
table(dat[,'HER2.Final.Status'])
```

5. How many women are ER-positive __and__ PR-positive? (Filter the dataset and then find the number of rows)
```{r}
dat1 <- dat[dat$ER.Status=='Positive' & dat$PR.Status=='Positive',] # You need the comma, since you're filtering on rows
nrow(dat1)
```
Alternatively,
```{r}
nrow(subset(dat, ER.Status=="Positive" & PR.Status=="Positive"))
```
Alternatively,
```{r}
sum(dat$ER.Status=='Positive' & dat$PR.Status=='Positive')
```

6. How many women are "triple-negative", i.e. negative for ER, PR and HER2?
```{r}
nrow(subset(dat, ER.Status=='Negative' & PR.Status=='Negative' & HER2.Final.Status=='Negative'))
```

Feel free to explore this data set further, to discover other characteristics. 
