---
title: "Practicum 8"
author: "Abhijit Dasgupta"
date: "October 31, 2017"
output: html_document
---

```{r setup, echo=F}
knitr::opts_chunk$set(echo = F, eval=F)

```

# Creating functions

## Standardizing continuous variables

1. Standardizing continuous variables to have mean 0 and variance 1 is a common practice
to help stabilize regression models. This involves subtracting the average value from each 
element of the variable, and then dividing the difference by the standard deviation. Write a 
function that takes a vector of numbers and standardizes it. Call it `standardize_mean`.
```{r}
standardize_mean <- function(x) {
  return((x - mean(x))/sd(x))
}

```

2. A more robust standardization that is often used to find outliers is the median standardization, where the
mean and standard deviation from 1 above are replaced by the median and median absolute deviation (`mad`), respectively. Create a `standardize_median` function.
```{r}
standardize_median <- function(x){
  return((x - median(x))/mad(x))
}
```


3. Update both functions to account for missing values
```{r}
standardize_mean <- function(x){
  return((x - mean(x, na.rm=T))/sd(x, na.rm=T))
}
```


4. Create a unified `standardize` function that takes an input `method` that will be either "mean" or "median", 
and output the corresponding standardized variable
```{r}
standardize <- function(x, method = 'mean'){
  if(method=='mean'){
    standardize_mean(x)
  }
  if(method == 'median'){
    standardize_median(x)
  }
}
```

## Metrics for binary outcomes

A lot of metrics for binary outcomes are derived from a contingency table, which is 
a 2 x 2 table of frequency counts comparing the true outcome with the predicted outcome. This can be
obtained using `table(truth, predicted)`, where `truth` and `predicted`  are provided by the user, and are
vectors of equal length comprising 1's and 0's. Using this table as input in each of the following functions:

1. Write a function for the  accuracy score, i.e. the proportion of cases in which the truth and predicted values are the same. This should be a fraction betwen 0 and 1.
2. Write a function (or functions) for computing sensitivity (true postive rate) and specificity (true negative rate)
3. Write a function for computing the positive predictive value, that is, the proprotion of predicted 1's that are actually true 1's, or in other words, P(true +ve | predicted +ve).
4. 

